---
title: "Data Cleaning"
author: "Robert McGuinn"
date: "2/4/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install, echo=F, warning=F, message=F}

library(tidyr)
library(dplyr)

```

# Read data

```{r}

##### read in data from URL #####

catch_orig <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method='libcurl'), stringsAsFactors = F)

reg_def <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = 'libcurl'), stringsAsFactors = F) %>% dplyr::select(code, mgmtArea) %>% rename(Region = code)

```

# Make data tidy

Remove the 'notes' and ''All' columns. They are not needed. Also removing problem value within Chinook where 'I' was written instead of '1'.  Fixing in this script below.

```{r}

##### filter ##### 

catch_clean <- catch_orig %>% 
  dplyr::select(-All, 
                -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
           mutate(Chinook = as.integer(Chinook)) 

##### checking #####

summary(catch_clean)

```

## Data QA

Checking for NA values in Chinook.

```{r}

##### find the places where Chinook went wrong #####
i <- which(is.na(catch_clean$Chinook))
i

##### look at this variable ##### 
catch_orig[i,]

```

Looks like 'l' was put in instead of '1'. 

## Convert to long format

We now want to convert to long format. Also making change to catch values, because the catch variable originally represented thousands of fish.  

```{r}

catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = 'species', 
               values_to = 'catch') %>%
  rename(catch_thousands = catch) %>% 
  mutate(catch = catch_thousands*1000) %>% 
  dplyr::select(-catch_thousands)
  
head(catch_long)

```

Reversing the transpose.

```{r}

catch_wide <- catch_long %>% 
  pivot_wider(names_from = 'species', values_from = 'catch')

head(catch_wide)
```


# Summarize

Here we calculate some means over each Region.

```{r}

mean_region <- catch_long %>% 
  group_by(Region ) %>% 
  summarise(mean = mean(catch), 
            n_obs=n())

```

Here we just look at 'Sockeye'

```{r}

options(scipen = 10000, digits = 3)

sockeye_catch <- catch_long %>% 
  filter(species == "Sockeye") %>% 
  group_by(Region) %>% 
  summarise(mean = mean(catch), 
            n_obs=n()) %>% 
  arrange(desc(mean))

```

# Join to other data

Merge region definitions. 

```{r}

catch_join <- left_join(catch_long, reg_def, 
                by = c("Region"))

##### checking ##### 

# head(catch_join)
# length(catch_join$Region)

```

This is how you to some parsing.


```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAI-039"),
                       stringsAsFactors = FALSE)

sites_df %>% 
  separate(site, into =  c("island", "site_number"), "-", remove = F)


```


Here is how you do a paste. 


```{r}
dates_df <- data.frame(year = c("1930",
                                "1930",
                                "1930"),
                       month = c("12",
                                "12",
                                "12"),
                       day = c("14",
                               "15",
                               "16"),
                       stringsAsFactors = FALSE)

dates_df %>% 
  unite(date, year, month, day, sep = "-", 
        remove = F)
```

